# Extra: CA1 {#sec-extraCA1 .unnumbered}

::: callout-note

<a href="downloads/CA1_extra_R_student.qmd" download="CA1_extra_R_student.qmd">
  <i class="bi bi-download"></i> Download file
</a> 

:::

## Overview {.unnumbered}

Coverage: L01-L09, P1-P3

This is a document to give you additional practice working in R. It will include working with data (manipulating, summarising, plotting) and analyzing the distributions we've covered.

Please do not treat this as a comprehensive study guide for CA1!

Note that there are often many possible paths to the correct solution.

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(tidyverse); theme_set(theme_classic())
library(readxl)
set.seed(2026)
```

## Iris

The `iris` dataset is available in R upon start-up.

What are the dimensions of the object (rows, columns) and what data type is it?

::: {.content-visible when-profile="key"}

::: callout-important

```{r}
#| code-fold: true
#| code-summary: "View code"

str(iris)
```
:::

:::

Create a scatter plot of sepal area on the x-axis and petal area on the y-axis.

::: {.content-visible when-profile="key"}

::: callout-important


First create new columns `Sepal.Area` and `Petal.Area`.

```{r}
#| message: false
#| code-fold: true
#| code-summary: "View code"

# method 1
iris_df <- iris |>
  mutate(Sepal.Area=Sepal.Length*Sepal.Width,
         Petal.Area=Petal.Length*Petal.Width)

# method 2
iris_df <- iris
iris$Sepal.Area <- iris$Sepal.Length * iris$Sepal.Width
iris$Petal.Area <- iris$Petal.Length * iris$Petal.Width
```

Then make the scatter plot:

```{r}
#| fig-width: 4
#| fig-height: 4
#| code-fold: true
#| code-summary: "View code"

plot(Petal.Area ~ Sepal.Area, data=iris)
```
:::

:::

For each species, what is the mean and standard deviation of sepal area?

::: {.content-visible when-profile="key"}

::: callout-important


```{r, message=FALSE}
#| code-fold: true
#| code-summary: "View code"

# method 1
iris |>
  group_by(Species) |>
  summarise(mean_SepalArea=mean(Sepal.Area),
            sd_SepalArea=sd(Sepal.Area))
```

```{r, eval=F}
#| code-fold: true
#| code-summary: "View code"

# method2
summary(iris$Species) # unique values of Species
mean(iris$Sepal.Area[iris$Species=="setosa"])
mean(iris$Sepal.Area[iris$Species=="versicolor"])
mean(iris$Sepal.Area[iris$Species=="virginica"])
sd(iris$Sepal.Area[iris$Species=="setosa"])
sd(iris$Sepal.Area[iris$Species=="versicolor"])
sd(iris$Sepal.Area[iris$Species=="virginica"])
```
:::

:::

Create a plot with two panels, each with a boxplot with species on the x-axis. In the first panel, plot sepal area on the y-axis. In the second panel, plot petal area on the y-axis.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

par(mfrow=c(1,2))
boxplot(Sepal.Area ~ Species, data=iris, cex.axis=0.7)
boxplot(Petal.Area ~ Species, data=iris, cex.axis=0.7)
```
:::

:::

Extract the even numbered rows from `iris`, and the columns `Species`, `Sepal.Area`, and `Petal.Area` in that order.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

iris[seq(2, nrow(iris), by=2), c("Species", "Sepal.Area", "Petal.Area")]
```
:::

:::

## Air quality

The `airquality` dataset is available in R upon start-up, and gives daily values of mean ozone, solar radiation, mean wind speed, and max temperature at LaGuardia Airport in New York City (`?airquality`).

Make a histogram of wind speed.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

hist(airquality$Wind)
```
:::

:::

Assume wind speed is normally distributed in this dataset with $\mu=9.99 mph$ and $\sigma = 3.51 mph$.

What proportion of days would you expect to have a wind speed less than 5 mph?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

pnorm(5, 9.99, 3.51)
```
:::

:::

You have a kite that can only be flown safely in winds between 14 mph and 20 mph. What proportion of days would be suitable?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

pnorm(20, 9.99, 3.51) - pnorm(14, 9.99, 3.51)
```
:::

:::

How many days would be suitable in a 31-day month?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

31 * (pnorm(20, 9.99, 3.51) - pnorm(14, 9.99, 3.51))
```
:::

:::

What proportion of days in the dataset were actually suitable?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

mean(airquality$Wind >= 14 & airquality$Wind <= 20)
```
:::

:::

What are the z-scores for 14 and 20 mph?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

(14 - 9.99)/3.51
(20 - 9.99)/3.51
```
:::

:::

What are the quantiles for the middle 95%?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

qnorm(c(0.025, 0.975), 9.99, 3.51)
```
:::

:::

What is the quantile for the largest 5%?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

qnorm(0.95, 9.99, 3.51)
```
:::

:::

Given the population parameters above, what is the z score of the sample mean in the dataset?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

(mean(airquality$Wind) - 9.99)/(3.51/sqrt(nrow(airquality)))
```
:::

:::

Given the population parameters above, what is the probability of observing a sample mean smaller than the one in this dataset?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

pnorm(mean(airquality$Wind), 9.99, 3.51/sqrt(nrow(airquality)))
```
:::

:::

Create a scatter plot with temperature on the x-axis and log-transformed ozone on the y-axis.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

plot(airquality$Temp, log(airquality$Ozone))
```
:::

:::

Create a boxplot with the solar radiation on the y-axis and the month on the x-axis.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

boxplot(Solar.R ~ Month, data=airquality)
```
:::

:::

## Coral predation

You are studying predation of the Flamingo Tongue Snail on soft corals on South Water Caye in Belize. In a distinct patch of reef, you randomly select a sample of 14 corals and identify that 9 of these have feeding scars.

What distribution would you expect this variable (i.e., the number of corals with feeding scars of those sampled) to follow?

::: {.content-visible when-profile="key"}

::: callout-important


> Binomial

:::

:::

What is your sample proportion and its standard error?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

p_hat <- 9/14
p_hat
sqrt((p_hat)*(1-p_hat)/(14-1))
```

:::

:::

Create a bar graph of the binomial distribution corresponding to these parameters, with the x-axis spanning the full range of possible outcomes.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

barplot(dbinom(0:14, 14, prob=9/14), names.arg=0:14)
```
:::

:::

Based on this distribution, if you were to collect another random sample of the same size, what is the probability that all 14 would have feeding scars?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

dbinom(14, 14, 9/14)
```
:::

:::

What is the probability you would find *fewer* with feeding scars than in the first sample?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

pbinom(8, 14, 9/14)
```
:::

:::

What is the probability of finding *more* with feeding scars than in the first sample?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

1-pbinom(9, 14, 9/14)
```
:::

:::

What is the probability of a sample with 4, 5, or 6 corals with feeding scars?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

sum(dbinom(4:6, 14, 9/14))
```
:::

:::

## Earthquakes

The dataset `quakes` is available in R upon start-up, and gives magnitudes and locations for 1000 seismic events near Fiji since 1964.

Plot the locations of the earthquakes (x-axis: longitude, y-axis: latitude).

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

plot(lat ~ long, data=quakes)
```
:::

:::

Create a column called `mag_cat` with the magnitude truncated to integer (i.e., 4.0-4.99 become 4). Hint: `?trunc`

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

quakes$mag_cat <- trunc(quakes$mag)
```
:::

:::

For each magnitude category (`mag_cat`), count the number of occurrences in the dataset.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

summary(quakes) # ranges 3-6
# Handy trick:
summary(as.factor(quakes$mag_cat))
```
:::

:::

Let's assume the dataset spans 60 years (1964-2023). What is the mean number of magnitude category 5 earthquakes per year?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

lambda_cat5 <- sum(quakes$mag_cat==5)/60
lambda_cat5
```
:::

:::

What distribution is most suitable for the number of category 5 earthquakes per year?

::: {.content-visible when-profile="key"}

::: callout-important


> Poisson

:::

:::

Using the above value as $\lambda$, create a bar plot of the probability distribution of the number of category 5 earthquakes per year, ranging from 0 to 10 earthquakes.

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

barplot(dpois(0:10, lambda_cat5), names.arg=0:10, xlab="Category 5 earthquakes per year")
```
:::

:::

What is the probability that next year has no category 5 earthquakes?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

dpois(0, lambda_cat5)
```
:::

:::

Of the next 10 years, how many would you expect to have:

No category 5 earthquakes?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

dpois(0, lambda_cat5)*10
```
:::

:::

Fewer than 2 earthquakes?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

ppois(1, lambda_cat5)*10
```
:::

:::

At least 5?

::: {.content-visible when-profile="key"}

::: callout-important


```{r}
#| code-fold: true
#| code-summary: "View code"

(1-ppois(4, lambda_cat5))*10
```
:::

:::
