# Visualization {#sec-P1}

Statistics can be divided into two broad categories: **descriptive statistics**, which describe or summarise data, and **inferential statistics**, which allow us to infer something about a population from a sample.

This session focuses on descriptive statistics: how to appropriately display and summarise. You already know several techniques for displaying data (e.g., bar charts, histograms, scatter plots). When and how to use these techniques is one focus of today.

There are two critical purposes of data visualization:

1.  Getting a 'gut' feel for your data. Examining data graphically enables you to identify any outliers (suspicious observations which could be errors). It will also help you to select the most appropriate inferential statistical model (more on this later).
2.  Communicating your results. Focus on clarity, drawing attention to the most interesting aspects of your data. Graphics that are confusing, either through a lack of detail (e.g. no labels) or that contain too much information will fail in this central objective.

As you create graphics, keep in mind that they may be viewed on different machines, in grey scale, or by colour-blind or visually impaired readers. Colour scales such as those available from [ColorBrewer](https://colorbrewer2.org) or [viridis](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) are designed with this in mind.

## R Review

We *very briefly* review key basics of R here. See **R Basics** for a more comprehensive review.

It's best practice to load necessary packages at the top of your document. Remember, packages need to be *installed once per machine*, but *loaded in each R session*.

To install a package, either click the link in the yellow banner at the top of your script (RStudio usually does this when needed), or run:

```{r}
#| eval: false

install.packages("tidyverse")
install.packages("cowplot")
```

**You will likely not need to do that again this semester!** However, you will need to use `library(packageName)` each time you open R.

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(readxl) # installed with tidyverse, but not loaded in library(tidyverse)
library(cowplot)
set.seed(2026) # for fully reproducible code
theme_set(theme_classic()) # plotting theme
```

------------------------------------------------------------------------

## Basic data exploration

### Object structure

We will mostly work with dataframes. A `data.frame` is a 2D rectangular object with columns and rows. In a tidy dataset, each row represents an 'observation' and each column represents a 'variable'. R (and often packages) contains several built-in dataframes, bundled into the `datasets` package which is automatically loaded.

```{r}
#| eval: false

# functions for basic details of objects
str(penguins) # structure overview
glimpse(penguins) # slightly different overview
names(penguins) # column names
head(penguins) # first few rows
```

Functions have *arguments*, which let you control their settings. They typically have defaults, which you can see in the help page. For example, `head()` shows the first 6 rows by default (`?head`: note `n = 6L`). Specify the argument to adjust the behaviour:

```{r}
head(penguins, n=2) 
tail(penguins, n=2)
# what are the last 10 rows?
```

::: callout-tip
Function arguments do not need to be named if they are listed in the order given in the help documentation. For example, `head(penguins, 2)` will do exactly the same thing as `head(x=penguins, n=2)`.
:::


### Subsetting, renaming, and rearranging

There are several ways to access subsets of a `data.frame`:

-   Use `data_df$columnName` to extract a single column
-   Use `data_df[rows,columns]` to extract a block

```{r}
#| eval: false

penguins$bill_len # whole column
```

```{r}
penguins[1, 3] # row 1, column 3
penguins[1:5, 3] # rows 1-5, column 3
penguins[2:4, ] # leaving the 'columns' blank will return all columns
```

We can also change column names. For illustration, let's make a copy of the `data.frame` to do that.

```{r}
penguins_alt <- penguins 
names(penguins_alt)
names(penguins_alt)[1] <- "species_name" # change first column name
names(penguins_alt)
```

Create a new column by specifying the name with `df$colName` and *assigning* a vector.
```{r}
penguins_alt$bill_area <- penguins_alt$bill_dep * penguins_alt$bill_len
head(penguins_alt, 2)
```

::: callout-note
Recall that you must *assign* the results of an operation (`<-`) to save it. For example, running `penguins_alt[, 2:1]` will display the results, but `penguins_alt <- penguins_alt[, 2:1]` will *overwrite* `penguins_alt` in your R environment.
:::

You can also subset based on criteria. Say we only want rows with observations from 2009:

```{r}
penguins_2009 <- penguins_alt[penguins_alt$year == 2009, ]
```

It is always good to check that you've done what you think you've done:

```{r}
nrow(penguins_2009) 
head(penguins_2009, 2)
count(penguins_alt, year)
count(penguins_2009, year)
```


### NAs and summary

When you import data, you should check for missing values. These are represented as `NA`.

We can check each element of a vector using `is.na()`, which will return `TRUE` if an element *is* `NA`, and `FALSE` if an element *is not* `NA`.

```{r}
#| eval: FALSE
#| echo: TRUE

is.na(penguins_alt$sex)
```

R converts logical values (i.e., `TRUE`/`FALSE`) to numeric (i.e., `1`/`0`) automatically, allowing convenient summaries:

```{r}
sum(is.na(penguins_alt$sex)) # how many are NA?
mean(is.na(penguins_alt$sex)) # what proportion are NA?
```

Another very useful check is `summary()`:

```{r}
summary(penguins_alt)
```

```{r}
#| eval: FALSE
summary(penguins_alt)
```

Once you are confident that your `data.frame` looks sensible, that it contains the data you expect, and that the data types are correct, you can start to explore and summarise your data.

There are many graphical methods for data exploration. The appropriate method depends on the nature of the data and what you wish to communicate to the reader.

------------------------------------------------------------------------

## Graphical methods for displaying data

Always keep in mind that **the primary reason for data visualization is to communicate concisely and accurately to your reader**. Graphics should be clear, concise and easy to understand. 

Real data is usually complex, and creating effective figures takes careful thought and often many revisions.


::: callout-note
We will primarily use the `ggplot2` package because it is more straightforward in many ways. It loads when you run `library(tidyverse)`. However, we will also note equivalent base R functions from time to time.
:::


### Scatter plots

Scatter plots display two continuous variables against each other. They are commonly used for analyses like correlation or linear regression. 

With `ggplot2`:

```{r}
#| fig-width: 4.5
#| fig-height: 4.5

ggplot(penguins_alt, aes(x=body_mass, y=bill_len)) + 
  geom_point() + 
  labs(x="Body mass (g)", 
       y="Bill length (mm)",
       title="Bill length increases with body mass")
```

The `ggplot()` function takes the dataframe as the first argument. The plot *aesthetics* are specified with `aes()`, such that variables (=columns) are *mapped* to plot aesthetics. Then, layers are added using `geom_*()`. Additional functions can be 'added' to further modify the plot. See a comprehensive list [in the ggplot2 manual](https://ggplot2.tidyverse.org/reference/index.html).

Once you understand the basic structure, it becomes easy to create more interesting figures:

```{r}
#| fig-width: 6
#| fig-height: 4.5

ggplot(penguins_alt, aes(x=body_mass, y=bill_len, colour=species_name)) + 
  geom_point() + 
  geom_line(stat="smooth", method="lm") +
  scale_colour_brewer("Species", palette="Dark2") +
  labs(x="Body mass (g)", 
       y="Bill length (mm)",
       title="Bill length increases with body mass")
```


With base R, the `plot()` function will create a scatter plot if given two numeric variables. There are two options for specifying the variables.

Using `data_df` with columns `x` and `y`, the same plot can be created with either syntax:

```{r}
#| eval: false

data_df <- data.frame(x=1:3, y=4:6)
plot(data_df$x, data_df$y) # two vectors: x, y
plot(y ~ x, data=data_df) # columns in dataframe: formula y ~ x
```

There are many options for modifying the output of `plot()`, but it is clunkier compared to `ggplot2`.

::: Q.
With the `penguins` dataset, plot bill length by body mass for only Chinstrap penguins. Make sure the plot is fully labelled.
:::

::: {.callout-tip collapse="true"}
## View solution

```{r}
#| eval: false

ggplot(penguins[penguins$species=="Chinstrap",], aes(body_mass, bill_len)) +
  geom_point() + 
  labs(x="Body mass (g)", 
       y="Bill length (mm)",
       title="Chinstrap penguins")
```
:::

```{r}
#| label: fig-chinstraps
#| fig-cap: "Chinstrap penguin allometrics."
#| fig-width: 6
#| fig-height: 4.5
#| echo: false

ggplot(penguins[penguins$species=="Chinstrap",], aes(body_mass, bill_len)) +
  geom_point() + 
  labs(x="Body mass (g)", 
       y="Bill length (mm)",
       title="Chinstrap penguins")
```


::: Q.
Using `ggplot()`, create a figure where each species is a different shape. See the **Aesthetics** section in `?geom_point` for help.
:::

```{r}
#| label: fig-penguinShapes
#| fig-cap: "Penguin allometrics."
#| fig-width: 6
#| fig-height: 4.5
#| echo: false

ggplot(penguins, aes(body_mass, bill_len, shape=species)) +
  geom_point() + 
  scale_shape_manual("Species", values=c(1, 2, 3)) +
  labs(x="Body mass (g)", 
       y="Bill length (mm)",
       title="Chinstrap penguins")
```


### Boxplots

Box plots are used to summarise a continuous variable by levels of a factor. 

Explore the `data.frame` using the strategies covered above. Which variables are categorical? Which are continuous?

```{r}
#| echo: false

str(penguins, 2)
```

```{r}
#| label: fig-billArea_spp
#| fig-cap: "Boxplot showing bill area by species"
#| fig-width: 4
#| fig-height: 4
#| echo: FALSE

ggplot(penguins_alt, aes(species_name, bill_area/body_mass, fill=species_name)) + 
  geom_boxplot() +
  labs(x="Species", 
       y="Bill area (mm^2) scaled by body mass (g)")
```

```{r}
#| eval: FALSE

?geom_boxplot
# See examples at bottom of the help page
```

::: Q.
Reproduce the plot shown in @fig-billArea_spp.
:::

::: {.callout-tip collapse="true"}
## View solution

```{r}
#| eval: false

ggplot(penguins_alt, aes(species_name, bill_area/body_mass, fill=species_name)) + 
  geom_boxplot() +
  labs(x="Species", 
       y="Bill area (mm^2) scaled by body mass (g)")
```
:::

::: Q.
Use `?boxplot.stats` to investigate what the box and whiskers actually represent.
:::

::: callout-note
In base R, use `boxplot(y ~ x, data=dataframe)` to create a boxplot.
:::

Note that box plots are not the most visually intuitive, since *smaller* box areas correspond with *more* observations. Packages like `ggplot2` (and extensions) make alternatives like those in @fig-billArea_spp_ggplot simple to produce. We will cover some of these later.

```{r}
#| label: fig-billArea_spp_ggplot
#| fig-cap: "Alternatives to boxplots. (A) Strip plot which shows all data along with means and 2 standard errors. (B) Rain cloud plot showing data distributions along with medians with middle 50% and 95%."
#| fig-width: 10
#| fig-height: 4
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

library(ggdist); library(cowplot)
theme_set(theme_classic())

p_a <- ggplot(penguins_alt, aes(species_name)) + 
  geom_jitter(aes(y=bill_area), alpha=0.25, width=0.05, size=1.5, shape=1) + 
  stat_summary(aes(y=bill_area), fun.data=~mean_se(.x, mult=2),
               geom="errorbar", width=0.2) +
  stat_summary(aes(y=bill_area), fun.data=~mean_se(.x, mult=2),
               geom="point", size=3) +
  labs(x="Species", y="Bill area (mm^2)")
p_b <- ggplot(penguins_alt, aes(bill_area, species_name)) + 
  geom_dots(side="bottom", scale=0.5) +
  stat_halfeye(scale=0.5, .width=c(0.5, 0.95)) + 
  labs(y="Species", x="Bill area (mm^2)")
plot_grid(p_a, p_b, align="hv", axis="tblr", nrow=1, labels="AUTO")
```

### Line plots

Line plots are most often seen in timeseries plots with time on the x-axis and the response on the y-axis. Line plots involve joining points with a line, which indicates that you have made assumptions about the value of the response variable between successive measurements.

We will examine these plots using the dataset `lynx`, which consists of the number of Canadian lynx pelts sold per year between 1821 - 1934. It is a 'classic' dataset as it shows a cyclical 'boom-and-bust' lynx population (demonstrating predator-prey interactions).

First, we will create a variable `Year`.

```{r}
str(lynx)
```

```{r}
#| results: "hide"

str(lynx)
lynx_df <- as.data.frame(lynx) 
str(lynx_df) 
lynx_df$Year <- seq(from=1821, to=1934, by=1)
lynx_df$Trappings <- as.numeric(lynx_df$x) # Time-Series is complicated.
str(lynx_df)
head(lynx_df, 2) 
```

In R, we use *functions* to perform actions on *objects*. Functions have arguments, taking the form `functionName(arg1=..., arg2=...)`. If you do not name the arguments, the function will assume that you are listing the arguments in order. See the help file for a function with `?` to see the argument order (e.g., `?seq`).

::: Q.
Using `seq()`, write a piece of code which generates odd numbers between 1 and 20. Try with and without naming the arguments.
:::

::: {.callout-tip collapse="true"}
## View solution

```{r}
#| eval: false
seq(from=1, to=20, by=2)
seq(1, 20, 2)
```
:::


::: Q.
Using `ggplot()`, produce a line plot similar to @fig-lynx_1.
:::

```{r}
#| label: fig-lynx_1
#| fig-cap: "The number of lynx trapped in Canada (1820-1934)"
#| echo: false
#| fig-align: 'center'

ggplot(lynx_df, aes(Year, Trappings)) +
  geom_point() +
  geom_line() + 
  labs(title="Lynx trapping in Canada (1820-1934)")
```

::: {.callout-tip collapse="true"}
## View solution

```{r}
#| eval: false
ggplot(lynx_df, aes(Year, Trappings)) +
  geom_point() +
  geom_line() + 
  labs(title="Lynx trapping in Canada (1820-1934)")
```
:::


::: callout-note
In base R, use `plot(y ~ x, data=dataframe, type='b')` to make plots with **b**oth points and lines, or `type='l'` for just **l**ines.
:::

::: Q.
Create a plot that shows the **log-transformed** number of trappings from **only** 1850 to 1900.
:::

::: {.callout-tip collapse="true"}
## View solution

```{r}
#| eval: false
ggplot(lynx_df[lynx_df$Year >= 1850 & lynx_df$Year <= 1900, ], 
       aes(Year, log(Trappings))) +
  geom_point() +
  geom_line() + 
  labs(title="Lynx trapping in Canada (1850-1900)")
```

Alternatively (and better, in this case) is to modify the *axis* rather than the variable. 

```{r}
ggplot(lynx_df[lynx_df$Year >= 1850 & lynx_df$Year <= 1900, ], 
       aes(Year, Trappings)) +
  geom_point() +
  geom_line() + 
  scale_y_log10() +
  guides(y="axis_logticks") +
  labs(title="Lynx trapping in Canada (1850-1900)")
```
:::

### Histograms

Histograms illustrate the distribution of **continuous** data. In histograms the bars are adjacent (no gap). This indicates that the underlying values are continuous rather than discrete.

```{r}
#| label: fig-lynx_hist_default
#| fig-cap: "Lynx pelts per year with default settings."
#| fig-width: 4
#| fig-height: 3

ggplot(lynx_df, aes(Trappings)) + # specify x-axis, bun not y-axis
  geom_histogram(colour="grey30", fill="grey90") + 
  labs(x="Trapped lynx per year", 
       title="Lynx trapping")
```

::::: Q.
What conclusions do you draw from this plot? Which range of values was most common across years?

:::: {.content-visible when-profile="key"}
::: callout-important
Low values (\< 1000) are most common, with a second mode between 2000-3000.
:::
::::
:::::

Be aware that histograms can be quite sensitive to the number of bins, and you should explore different options. You can set the number of bins with `bins=...` or specific values of break points with `breaks=...`.

::: callout-note
To make a multi-panel figure with ggplot, use the `plot_grid()` function in the `cowplot` package.
:::

```{r}
#| label: fig-lynx_hist_b1
#| fig-cap: "Lynx pelts per year with breaks=5 (left) and a vector of breaks (right)."
#| fig-width: 6
#| fig-height: 3

library(cowplot)
pA <- ggplot(lynx_df, aes(Trappings)) +
  geom_histogram(colour="grey30", fill="grey90", 
                 bins=5) + 
  labs(x="Trapped lynx per year", 
       title="Lynx trapping")
pB <- ggplot(lynx_df, aes(Trappings)) +
  geom_histogram(colour="grey30", fill="grey90", 
                 breaks=c(0, 500, 1000, 2000, 5000, 10000)) + 
  labs(x="Trapped lynx per year", 
       title="Lynx trapping")
plot_grid(pA, pB, nrow=1, labels="AUTO")
```

```{r}
#| label: fig-lynx_hist_panels
#| fig-cap: "Histograms of lynx pelts per year with breaks every (A) 100, (B) 500, (C) 1,000, or (D) 2,000."
#| fig-width: 6
#| fig-height: 4
#| code-fold: true

pA <- ggplot(lynx_df, aes(Trappings)) +
  geom_histogram(colour="grey30", fill="grey90", 
                 breaks=seq(0, 10000, by=100)) + 
  labs(x="Trapped lynx per year")
pB <- ggplot(lynx_df, aes(Trappings)) +
  geom_histogram(colour="grey30", fill="grey90", 
                 breaks=seq(0, 10000, by=500)) + 
  labs(x="Trapped lynx per year")
pC <- ggplot(lynx_df, aes(Trappings)) +
  geom_histogram(colour="grey30", fill="grey90", 
                 breaks=seq(0, 10000, by=1000)) + 
  labs(x="Trapped lynx per year")
pD <- ggplot(lynx_df, aes(Trappings)) +
  geom_histogram(colour="grey30", fill="grey90", 
                 breaks=seq(0, 10000, by=2000)) + 
  labs(x="Trapped lynx per year")
plot_grid(pA, pB, pC, pD, nrow=2, ncol=2,
          align="hv", axis="tblr", labels="AUTO")
```


Which of these plots is the most useful? There is no definitive answer, but the first is very busy and the last fails to show relevant detail near 0. Bin widths of 500-1000 communicate the patterns most clearly.

Generally, 5-15 breaks usually work well.


::: callout-note
In base R, use `hist(dataframe$column)`.
:::

### Bar graphs

Bar graphs are used to plot counts of categorical or discrete variables. We'll be using the `islands` dataset which is a named vector of island areas.

::: callout-note
Many objects in R can have row names. However, converting between data types may lose this information. Consequently, **it is better practice to store relevant information in a column**. Nevertheless, there are occasions where this is useful and you may come across datasets with data stored as row names.
:::

Working with data involves a lot of time spent tidying the datasets: cleaning, checking, and reshaping into useful formats. We will cover a more modern set of methods for this later in the course using the *tidyverse* package. For now, we'll stay with base R. First, we need to tidy the `islands` data.

```{r}
#| results: "hide"

str(islands) # this is a named numeric vector
head(islands)

# convert to a dataframe
islands_df <- as.data.frame(islands) 
head(islands_df, 2)
str(islands_df) # rownames are not shown!
```

```{r}
# put the row names into a new column
islands_df$LandMass <- row.names(islands_df) 
head(islands_df, 2)

# set row names to the row number
row.names(islands_df) <- 1:nrow(islands_df) 
names(islands_df)[1] <- "Area" 
head(islands_df, 2) 
```

An initial attempt at a barplot makes a terrible figure.

```{r}
#| label: fig-island_1
#| fig-cap: "Island areas with barplot defaults"
#| fig-width: 6
#| fig-height: 2.5

ggplot(islands_df, aes(LandMass, Area)) + 
  geom_bar(stat="identity") # 'identity' means it won't calculate any summaries
```

The whole dataset includes a lot of very small areas, so let's cut it down to just the 10 largest. We also want to order the land masses by area in our plot. 

We will use the `arrange()` function to order our dataframe, and then use `scale_x_discrete()` to specify the breaks in the desired order.

```{r}
islands_df <- arrange(islands_df, desc(Area)) # desc() makes it largest to smallest
```


```{r}
#| label: fig-island_2
#| fig-cap: "Top 10 island areas"
#| fig-width: 4
#| fig-height: 4

ggplot(islands_df[1:10,], aes(LandMass, Area)) + 
  geom_bar(stat="identity") +
  scale_x_discrete(limits=unique(islands_df$LandMass)[1:10])
```

It would be more effective sideways:

```{r}
#| label: fig-island_3
#| fig-cap: "Finally! Did you know Antarctica is bigger than Europe?"
#| fig-width: 4
#| fig-height: 4


ggplot(islands_df[1:10,], aes(Area, LandMass)) + 
  geom_bar(stat="identity") +
  scale_y_discrete(limits=unique(islands_df$LandMass)[10:1]) +
  labs(x="Land mass", 
       y="Area (km^2)")
```

Data visualization is an iterative process with lots of trial and error to find a plot that communicates the message within the data well. 


------------------------------------------------------------------------

## Summary statistics

You will often need to summarise your data before you present it. Data summaries are usually contained in tables and they can sometimes replace graphics (e.g., where the data is relatively simple or where individual precise values are important). There are many types of summary statistics. Here we are concerned with central tendency and variability.

::::: Q.
What are the three main measures of central tendency?

:::: {.content-visible when-profile="key"}
::: callout-important
Mean, median, and mode
:::
::::
:::::

::::: Q.
What are three measures of variability?

:::: {.content-visible when-profile="key"}
::: callout-important
Range, interquartile range (IQR), and variance / standard deviation
:::
::::
:::::

The most appropriate metrics of central tendency or variability will depend on your data. Another summary statistic that you might include is sample size. R is very good at producing summary statistics, and there are myriad ways to produce them. 

Often you'll wish to summarise your data across levels of a certain factor, such as levels of a certain treatment. More complex summaries can be made using the *dplyr* package. We'll go into more detail later on some of the very powerful ways this package (and others in the *tidyverse*) can be used.

We'll use the built-in dataset `InsectSprays`. Viewing your raw data can be an important check as well. You can open a spreadsheet-style viewer in R using `View(YourDataFrame)` or by clicking on the icon in the *Environment* pane.

```{r}
str(InsectSprays)
```

To do more complex summaries, we will string together a series of functions. This can be done in a nested format (e.g., `fun3(fun2(fun1(dataset)))`), but this gets unwieldy very quickly.

So, let's use the *pipe* operator `|>`. This takes the output from one function and feeds it as the first input of the next (e.g., `dataset |> fun1() |> fun2() |> fun3()`), making code much more legible. Many functions in the *tidyverse* are built for piping.

```{r}
#| eval: FALSE

?`|>`
```

```{r}
#| results: "hide"

# use group_by() with the grouping column name(s)
spray_summaries <- InsectSprays |>
  group_by(spray) |>
  summarise(count_mean=mean(count))
spray_summaries
```

```{r}
# it is very easy to calculate any number of summary statistics
InsectSprays |>
  group_by(spray) |>
  summarise(mean=mean(count) |> signif(3),
            median=median(count),
            max=max(count),
            sd=sd(count) |> signif(3),
            N=n(),
            N_over_10=sum(count > 10),
            Pr_over_5=mean(count > 5))
```

### Choosing a central tendency metric

The choice of central tendency metric depends on the nature of the data and objectives of your research. We will use datasets that you downloaded from Brightspace (Practicals \> data). Remember to put these into the *data* folder in your working directory (or modify the file paths in the code accordingly).

```{r}
# this will load the 'Scallop %fat' data sheet from the xlsx spreadsheet.
scallop_df <- read_excel("data/H2SMS_practicalData.xlsx", sheet="Scallop %fat")
str(scallop_df)
# avoid spaces and symbols in column names. It's a pain.
names(scallop_df) <- "fat_pct"
```

::::: Q.
Check the data using the methods above. Does it look OK to you?

:::: {.content-visible when-profile="key"}
::: callout-important
There is an outlier with value 99.00
:::
::::
:::::

::::: Q.
Are these data likely to be continuous or discontinuous?

:::: {.content-visible when-profile="key"}
::: callout-important
Continuous
:::
::::
:::::

::::: Q.
Create a plot to visualize the distribution of these data.

:::: {.content-visible when-profile="key"}
::: callout-important
```{r}
ggplot(scallop_df, aes(fat_pct)) +
  geom_histogram(bins=15)
```
:::
::::
:::::

::::: Q.
Do you spot any issues?

:::: {.content-visible when-profile="key"}
::: callout-important
As noted above, there is a clear outlier.
:::
::::
:::::

::: {.callout-tip collapse="true"}
## View solution

```{r}
#| label: fig-scallop_hist
#| fig-cap: "Histogram of fat percentage."
#| fig-width: 4
#| fig-height: 4

ggplot(scallop_df, aes(fat_pct)) +
  geom_histogram(bins=15)
```
:::

You should have spotted a potential outlier. Data entry errors are common, and a check against the original data sheet shows that the decimal was typed in the wrong place. The following code helps you locate the error.

```{r}
which(scallop_df$fat_pct > 50) # which() returns the indexes
scallop_df$fat_pct[35:37] # row 36 is 99.0, but should be 9.90
scallop_df <- scallop_df[, c(1, 1)] # duplicate column
names(scallop_df) <- c("fat_pct_orig", "fat_pct_corr")
head(scallop_df, 2)
```

```{r}
#| results: "hide"

# there are many ways to 'fix' the outlier in R.
# You need to correct the outlier in row 36 of column 'fat_pct_corr'
scallop_df$fat_pct_corr[36] <- 9.9
which(scallop_df$fat_pct_corr > 90) 
# integer(0) - this means that no elements in fat_pct_corr contain values >90
```

Now summarise `scallop_df` using some of the methods above.

::::: Q.
Create a histogram for the corrected column. How does it differ from the original column with the error?

:::: {.content-visible when-profile="key"}
::: callout-important
The outlier is gone.

```{r}
ggplot(scallop_df, aes(fat_pct_corr)) +
  geom_histogram(bins=15)
```
:::
::::
:::::

::::: Q.
Calculate mean, variance, median, interquartile range, minimum, maximum and range for both `fat_pct_orig` and `fat_pct_corr`.

:::: {.content-visible when-profile="key"}
::: callout-important
Could do this line by line, or modify the code above:

```{r}
scallop_df |>
  summarise(mean=mean(fat_pct_orig),
            var=var(fat_pct_orig),
            med=median(fat_pct_orig),
            iqr=IQR(fat_pct_orig),
            min=min(fat_pct_orig),
            max=max(fat_pct_orig),
            rng=diff(range(fat_pct_orig)))

scallop_df |>
  summarise(mean=mean(fat_pct_corr),
            var=var(fat_pct_corr),
            med=median(fat_pct_corr),
            iqr=IQR(fat_pct_corr),
            min=min(fat_pct_corr),
            max=max(fat_pct_corr),
            rng=diff(range(fat_pct_corr)))
```
:::
::::
:::::

::::: Q.
Suppose the outlier was even bigger (i.e. your typo was even worse). Adjust your data, multiplying the erroneous data item by 10; copy the `fat_pct_orig` column and change row 36 to 999.

:::: {.content-visible when-profile="key"}
::: callout-important
```{r}
scallop_df$fat_pct_badTypo <- scallop_df$fat_pct_orig
scallop_df$fat_pct_badTypo[36] <- 999
```
:::
::::
:::::

::::: Q.
Calculate the same summary statistics.

:::: {.content-visible when-profile="key"}
::: callout-important
```{r}
scallop_df |>
  summarise(mean=mean(fat_pct_badTypo),
            var=var(fat_pct_badTypo),
            med=median(fat_pct_badTypo),
            iqr=IQR(fat_pct_badTypo),
            min=min(fat_pct_badTypo),
            max=max(fat_pct_badTypo),
            rng=diff(range(fat_pct_badTypo)))
```
:::
::::
:::::

::::: Q.
Which measures of central tendency and variability are most 'robust' against this outlier?

:::: {.content-visible when-profile="key"}
::: callout-important
Median and IQR, since they are based on ranks.
:::
::::
:::::

Or look individually instead of calculating many metrics at once with `tidyverse` functions:

```{r}
summary(scallop_df$fat_pct_corr)
var(scallop_df$fat_pct_corr)
IQR(scallop_df$fat_pct_corr)
```

R is excellent at generating well formatted tables such as shown in @tbl-scallop. What is missing from from this table?

```{r}
#| label: tbl-scallop
#| tbl-cap: "Summary statistics with and without an outlier. Note which summary stats are most influenced by the outlier."
#| echo: false

scallop_df |>
  pivot_longer(1:2, names_to="Column", values_to="val") |>
  group_by(Column) |>
  summarise(Mean=mean(val) |> signif(3),
            Median=median(val) |> signif(3),
            `Standard deviation`=sd(val) |> signif(3),
            Range=range(val) |> diff() |> signif(3),
            `Interquartile range`=IQR(val) |> signif(3)) |>
  knitr::kable(booktabs=TRUE, digits=3)
```

::::: Q.
How would the patterns seen in @tbl-scallop influence your choice if you were required to summarise data that you thought might contain values that could be erroneous? Consider how each metric is influenced by the data distribution and by outliers.

:::: {.content-visible when-profile="key"}
::: callout-important
Median and IQR are most robust to entry-related errors.
:::
::::
:::::

Let's load a dataset that gives the length of hake across three years of sampling.

```{r}
hake_df <- read_excel("data/H2SMS_practicalData.xlsx", sheet="Hake")
str(hake_df) # once again, column names made for excel rather than R
```

::::: Q.
What type of variable is `length`?

:::: {.content-visible when-profile="key"}
::: callout-important
Numeric, continuous
:::
::::
:::::

::::: Q.
Select an appropriate graphical method and display these data.

::: callout-tip
Hint: `ggplot(..., aes(..., group=Year)) + geom_boxplot()`
:::

:::: {.content-visible when-profile="key"}
::: callout-important
```{r}
#| fig-width: 9
#| fig-height: 4

pA <- ggplot(hake_df, aes(Year, `Hake length (mm)`)) + 
  geom_point(alpha=0.25, shape=1) # alpha sets the opacity
pB <- ggplot(hake_df, aes(Year, `Hake length (mm)`, group=Year)) + 
  geom_boxplot()
plot_grid(pA, pB, nrow=1)
```
:::
::::
:::::

::::: Q.
In your own time, summarise the hake data by year using the *tidyverse* functions you used above with the `InsectSprays` data.

:::: {.content-visible when-profile="key"}
::: callout-important
```{r}
hake_df |>
  group_by(Year) |>
  summarise(mean=mean(`Hake length (mm)`),
            var=var(`Hake length (mm)`),
            med=median(`Hake length (mm)`),
            iqr=IQR(`Hake length (mm)`),
            min=min(`Hake length (mm)`),
            max=max(`Hake length (mm)`),
            rng=diff(range(`Hake length (mm)`)))
```
:::
::::
:::::

```{r}
hake_df$Year <- as.factor(hake_df$Year) # Treat as categorical, not numeric
names(hake_df) <- c("Year", "Length") # simplify the column names
```

```{r}
#| label: tbl-hake
#| tbl-cap: "Summary of hake data."
#| echo: false

hake_df |>
  group_by(Year) |>
  summarise(`Mean length (cm)`=mean(Length)) |>
  knitr::kable(booktabs=TRUE, digits=1)
```

The following 'settling velocity' data relates to the settling velocity of salmon faecal material. Shona Magill generated these data.

```{r}
fishPoo_df <- read_excel("data/H2SMS_practicalData.xlsx", sheet="Settling velocity")
str(fishPoo_df)
```

::::: Q.
Produce a histogram of the settling velocity. Is it left or right skewed?

:::: {.content-visible when-profile="key"}
::: callout-important
It is right skewed.

```{r}
ggplot(fishPoo_df, aes(`Settling velocity (mm s-1)`)) + geom_histogram()
```
:::
::::
:::::

::::: Q.
Which measures of central tendency and variability are most appropriate?

:::: {.content-visible when-profile="key"}
::: callout-important
Median and IQR, most likely.
:::
::::
:::::

::::: Q.
Considering the distribution, is the mean or median larger?

:::: {.content-visible when-profile="key"}
::: callout-important
The mean is larger.
:::
::::
:::::

::::: Q.
Generate a new column of the log-transformed settling velocity data and plot these data.

:::: {.content-visible when-profile="key"}
::: callout-important
```{r}
ggplot(fishPoo_df, aes(`Settling velocity (mm s-1)`)) + 
  geom_histogram() + 
  scale_x_log10() +
  guides(x="axis_logticks")
```
:::
::::
:::::

::::: Q.
What measures of central tendency and variability could be applied to the log-transformed data? Selecting the preferable metrics for a dataset is not necessarily straightforward.

:::: {.content-visible when-profile="key"}
::: callout-important
Any, but mean and variance / standard deviation would be best.
:::
::::
:::::

@tbl-metric gives some indication of what issues you might consider.

```{r}
#| label: tbl-metric
#| tbl-cap: "Appropriate measures of central tendency and variability according to the underlying data distribution."
#| echo: false

tibble(a=c("Continuous, unimodal, symmetric", "Continuous, skewed", 
           "Continuous, multimodal", "Discontinuous"),
       b=c("Mean", "Median", 
           "None; state modes", "None; data-dependent"),
       c=c("Variance or sd", "Interquartile range", 
           "None; summarise by group", "Range?")) |>
  rename(`Data distribution`=a, 
         `Central tendency metric`=b,
         `Variability metric`=c) |>
  knitr::kable(booktabs=T) 
```

------------------------------------------------------------------------

## Conclusions

Visualizing and summarising data are the critical first steps in the data analysis and reporting workflow. We use graphical methods to firstly explore our own data. Once we have made sense of it we select the most appropriate method to convey that understanding to our readers. We may help that communication by summarising data in the most appropriate way taking into account the distribution of the data and the presence of outliers.
