# Problem Set 5 {#sec-P5_probSet .unnumbered}

::: callout-note

<a href="downloads/probSets_P5_student.qmd" download="probSets_P5_student.qmd">
  <i class="bi bi-download"></i>  Download file
</a> 

Please render this file to .html and submit via Brightspace.
:::

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(tidyverse); theme_set(theme_classic())
library(readxl)
library(emmeans)
set.seed(2026)
```

## T-tests, ANOVA, and regression

For each scenario, select an appropriate analysis given the dataset. Follow the workflow illustrated in the practical and in the lectures, checking assumptions and applying transformations as needed.

### Scenario 1


**Research question:** Do critical thermal limits change with repeated heatwave exposure in kelp?

In an experiment similar to [King et al. 2024](https://doi.org/10.1111/1365-2435.14622), a (hypothetical) H4 student measured critical thermal maxima (CT~max~) on the kelp species *Laminaria hyperborea* under 1) no previous heat exposure, 2) recent moderate heat exposure, and 3) recent extreme heat exposure. 

Identify the response and predictor variables, as well as what type of variable each is. 

>

::: {.content-hidden when-profile="man"}
::: callout-important

CT~max~ is the response; it is continuous.

Heat exposure is the predictor; it is categorical.

:::
:::

Next, load the data from the "MHW" sheet in "H2SMS_practicalData.xlsx" and perform appropriate checks.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important
```{r}
#| code-fold: true
#| code-summary: "View code"

mhw_df <- read_xlsx("data/H2SMS_practicalData.xlsx", sheet="MHW")
str(mhw_df)
summary(mhw_df)
mhw_df$treatment <- factor(mhw_df$treatment, levels=c("Control", "Moderate", "Extreme"))
```

:::
:::

Which analysis is most appropriate?

>

::: {.content-hidden when-profile="man"}
::: callout-important

ANOVA

:::
:::


What is the null hypothesis?

>

::: {.content-hidden when-profile="man"}
::: callout-important

$\mu_{Control} = \mu_{Moderate} = \mu_{Extreme}$

:::
:::


Perform the analysis, including assumption checks. Calculate appropriate confidence intervals, make appropriate comparisons, and summarise results concisely in one short paragraph including key statistics.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important

```{r}
#| code-fold: true
#| code-summary: "View code"

mhw_aov <- aov(CTmax ~ treatment, data=mhw_df)
```

```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 7
#| fig-height: 6

par(mfrow=c(2,2))
plot(mhw_aov)
par(mfrow=c(1,1))
```

Assumptions look reasonable. Approximately normally distributed residuals and similar residual variance in each treatment.

```{r}
#| code-fold: true
#| code-summary: "View code"

summary(mhw_aov)
```

There is a significant effect of treatment (p < 0.05).

```{r}
#| code-fold: true
#| code-summary: "View code"

em_df <- emmeans(mhw_aov, specs="treatment")
em_df
TukeyHSD(mhw_aov)
```

> There was a significant difference in CT~max~ among all treatments (F~2,87~=5502, p < 0.001). The individuals in the control showed the most heat tolerance (mean CT~max~ [95% CI]: 30.1 C [30.0-30.14 C]), with a slight decline in individuals previously exposed to moderate heatwave conditions (29.0 C [28.9-29.1 C]) and a dramatic decrease in heat tolerance for individuals previously exposed to extreme heatwave conditions (24.0 C [23.9-24.1 C]). 

::: callout-note
Depending on the specific question and focus, you might focus on the mean densities (listing them specifically in text) or on the differences: use the text to emphasize the main results of interest. Further detail can be included in figures and tables.
:::


```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-cap: "CT~max~ in *Laminaria hyperborea* under different prior heatwave exposures. Observed values (small points) along with mean (open points) and 95% confidence intervals."
#| fig-width: 3.5
#| fig-height: 3.5

ggplot(mhw_df, aes(treatment)) +
  geom_point(aes(y=CTmax), alpha=0.5) +
  geom_point(data=as_tibble(em_df), aes(y=emmean), shape=1, size=3) +
  geom_errorbar(data=as_tibble(em_df), aes(ymin=lower.CL, ymax=upper.CL), width=0.25) + 
  labs(x="Treatment", 
       y="CTmax (C)")
```

:::
:::




### Scenario 2

**Research question:** Is the average concentration of *Alexandrium spp* in an area greater than 40 cells per litre?

In the UK, shellfish cannot be harvested on aquaculture farms if densities of harmful algal species exceed certain levels. For example, *Alexandrium spp* release paralytic shellfish toxins, and growers are not allowed to harvest when *Alexandrium spp* exceeds 40 cells per litre. 

I've collected 13 water samples and enumerated the number of *Alexandrium spp* cells. 

::: callout-note
NB: This is *not* quite how the regulatory monitoring works!
:::

Identify the response and predictor variables, as well as what type of variable each is. 

>

::: {.content-hidden when-profile="man"}
::: callout-important

Alexandrium density is the response; it is continuous.

There is not a predictor variable.

:::
:::

Next, load the data from the "Alexandrium" sheet in "H2SMS_practicalData.xlsx" and perform appropriate checks.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important
```{r}
#| code-fold: true
#| code-summary: "View code"

alex_df <- read_xlsx("data/H2SMS_practicalData.xlsx", sheet="Alexandrium")
str(alex_df)
summary(alex_df)
```

:::
:::

Which analysis is most appropriate?

>

::: {.content-hidden when-profile="man"}
::: callout-important

One-sample t-test

:::
:::

What is the null hypothesis?

::: {.content-hidden when-profile="man"}
::: callout-important

$\mu_{Alex} \leq 40$

:::
:::

Perform the analysis, including assumption checks. Calculate appropriate confidence intervals, make appropriate comparisons, and summarise results concisely in one short paragraph including key statistics.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important

```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 9
#| fig-height: 4

par(mfrow=c(1,2))
hist(alex_df$alex_cells_L, breaks=5)
qqnorm(alex_df$alex_cells_L); qqline(alex_df$alex_cells_L)
```

Data are reasonably normal.

```{r}
#| code-fold: true
#| code-summary: "View code"

t.test(alex_df$alex_cells_L, mu=40, alternative="greater")
```

> We did not find evidence that the mean density of *Alexandrium* cells was above 40 cells L^-1^ (t~12~=-11.1, p = 1), with an observed mean density of 30.4 (one-tailed 95% confidence limit: 28.8).

:::
:::


### Scenario 3


**Research question:** Does mean seagrass shoot density differ among patches?

The data consists of seagrass shoot density measured by snorkelers in 20 randomly located quadrats in each of 3 lochs in western Scotland. 

Identify the response and predictor variables, as well as what type of variable each is. 

>

::: {.content-hidden when-profile="man"}
::: callout-important

Seagrass shoot density is the response; it is continuous.

Loch is the predictor; it is categorical.

:::
:::

Next, load the data from the "Seagrass" sheet in "H2SMS_practicalData.xlsx" and perform appropriate checks.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important
```{r}
#| code-fold: true
#| code-summary: "View code"

seagrass_df <- read_xlsx("data/H2SMS_practicalData.xlsx", sheet="Seagrass")
str(seagrass_df)
summary(seagrass_df)
seagrass_df$loch <- factor(seagrass_df$loch)
```

:::
:::

Which analysis is most appropriate?

>

::: {.content-hidden when-profile="man"}
::: callout-important

ANOVA

:::
:::


What is the null hypothesis?

>

::: {.content-hidden when-profile="man"}
::: callout-important

$\mu_{Ailort} = \mu_{Duich} = \mu_{Hourn}$

:::
:::


Perform the analysis, including assumption checks. Calculate appropriate confidence intervals, make appropriate comparisons, and summarise results concisely in one short paragraph including key statistics.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important

```{r}
#| code-fold: true
#| code-summary: "View code"

seagrass_aov <- aov(obs ~ loch, data=seagrass_df)
```

```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 7
#| fig-height: 6

par(mfrow=c(2,2))
plot(seagrass_aov)
par(mfrow=c(1,1))
```

Normality is not good, and the variance seems to be different among lochs, with an apparent increase with the fitted value (i.e., variance increases with the mean). A log transform is a reasonable choice to try.

```{r}
#| code-fold: true
#| code-summary: "View code"

seagrass_df$log_obs <- log(seagrass_df$obs)
seagrass_aov <- aov(log_obs ~ loch, data=seagrass_df)
```

```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 7
#| fig-height: 6

par(mfrow=c(2,2))
plot(seagrass_aov)
par(mfrow=c(1,1))
```

Not perfect, but much better and close enough.

```{r}
#| code-fold: true
#| code-summary: "View code"

summary(seagrass_aov)
```

There is no significant effect of loch (p > 0.05).

> There was no a significant difference in mean seagrass shoot density among lochs (F~2,57~=0.906, p = 0.41).  

```{r}
#| code-fold: true
#| code-summary: "View code"

em_df <- emmeans(seagrass_aov, specs="loch")
em_df <- as_tibble(em_df) |>
  mutate(exp_mn=exp(emmean),
         exp_lo=exp(lower.CL),
         exp_hi=exp(upper.CL))
```


```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-cap: "Seagrass shoot density. Observed values (small points) along with mean (open points) and 95% confidence intervals."
#| fig-width: 3.5
#| fig-height: 3.5

ggplot(seagrass_df, aes(loch)) +
  geom_point(aes(y=obs), alpha=0.5) +
  geom_point(data=em_df, aes(y=exp_mn), shape=1, size=3) +
  geom_errorbar(data=em_df, aes(ymin=exp_lo, ymax=exp_hi), width=0.25) + 
  labs(x="Loch", 
       y="Seagrass shoot density (n/m2)")
```

Or, potentially consider the log-scale version:
```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-cap: "Seagrass shoot density. Observed values (small points) along with mean (open points) and 95% confidence intervals."
#| #| fig-width: 3.5
#| fig-height: 3.5

ggplot(seagrass_df, aes(loch)) +
  geom_point(aes(y=obs), alpha=0.5) +
  geom_point(data=em_df, aes(y=exp_mn), shape=1, size=3) +
  geom_errorbar(data=em_df, aes(ymin=exp_lo, ymax=exp_hi), width=0.25) + 
  labs(x="Loch", 
       y="Seagrass shoot density (n/m2)") +
  scale_y_log10() + 
   guides(y="axis_logticks")
```


:::
:::




### Scenario 4

**Research question:** Does egg hatching success rate of *Acartia tonsa* depend on temperature?

A (hypothetical) H4 student has decided to replicate one of the experiments of [Holste and Peck (2006)]( https://doi.org/10.1007/s00227-005-0132-0) to test the effect of temperature on egg hatching success of the copepod *Acartia tonsa*. Their data are in the "HatchTemp" sheet in "H2SMS_practicalData.xlsx", containing the temperature (C) and hatching success rate (%) for 3 replicates at each temperature.


```{r}
#| include: false
#| eval: false

calc_hatch_prob <- function(temperature) {
  a <- rnorm(1, 5.73, 7.68)
  b <- rnorm(1, 3.8, 0.44)
  return(a + b*temperature)
}

set.seed(1)
HS_df <- expand_grid(temp=8:18,
                     replicate=1:3) |>
  rowwise() |>
  mutate(HS=calc_hatch_prob(temp)) |> 
  ungroup() |>
  mutate(obs=HS + rnorm(n(), 0, 3))
if(FALSE) {
  write_csv(HS_df, "temp_hatchTemp.csv")
}
HS_df |>
  ggplot(aes(temp, obs)) + geom_point() + ylim(0, 100) + stat_smooth(method="lm")
par(mfrow=c(2,2))
plot(lm(obs ~ temp, data=HS_df)) 
par(mfrow=c(1,1))
summary(lm(obs ~ temp, data=HS_df))

```


Identify the response and predictor variables, as well as what type of variable each is. 

>

::: {.content-hidden when-profile="man"}
::: callout-important

Hatching success rate is the response; it is continuous and constrained 0-100.

Temperature is the predictor; it is continuous

:::
:::

Next, load the data from the "HatchTemp" sheet in "H2SMS_practicalData.xlsx" and perform appropriate checks.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important
```{r}
#| code-fold: true
#| code-summary: "View code"

hatchTemp_df <- read_xlsx("data/H2SMS_practicalData.xlsx", sheet="HatchTemp")
str(hatchTemp_df)
summary(hatchTemp_df)
```

:::
:::

Which analysis is most appropriate?

>

::: {.content-hidden when-profile="man"}
::: callout-important

Linear regression

:::
:::


What is the null hypothesis?

>

::: {.content-hidden when-profile="man"}
::: callout-important

$\beta = 0$

:::
:::


Perform the analysis, including assumption checks. Calculate appropriate confidence intervals, make appropriate comparisons, and summarise results concisely in one short paragraph including key statistics.

```{r}

```

::: {.content-hidden when-profile="man"}
::: callout-important

```{r}
#| code-fold: true
#| code-summary: "View code"

hatchTemp_lm <- lm(hatch_percent ~ temp, data=hatchTemp_df)
```

```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 7
#| fig-height: 6

par(mfrow=c(2,2))
plot(hatchTemp_lm)
par(mfrow=c(1,1))
```

Assumptions look reasonable. Approximately normally distributed residuals and similar residual variance in across temperatures.

::: callout-note
Note that the response variable is a percentage. In this case, the assumptions look reasonable for the original data, but be aware that extrapolating to temperatures beyond the measured range would very quickly lead to impossible predictions! 
:::

```{r}
#| code-fold: true
#| code-summary: "View code"

summary(hatchTemp_lm)
```

There is a significant effect of temperature (p < 0.05).

```{r}
#| code-fold: true
#| code-summary: "View code"

confint(hatchTemp_lm)
```

> Hatching success rate increased with temperature (mean [95% CI]: 4.17 [3.1-5.24] percentage points per degree C; p < 0.001, R^2^~adj~ = 0.66, n=33). 

```{r}
#| code-fold: true
#| code-summary: "View code"
#| fig-cap: "Hatching success rate by temperature. Observed values (small points) along with fitted line and 95% confidence interval."
#| fig-width: 3.5
#| fig-height: 3.5

ggplot(hatchTemp_df, aes(temp, hatch_percent)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm") +
  scale_x_continuous("Temperature (C)", breaks=seq(8, 18, by=2)) +
  scale_y_continuous("Hatching success (%)", limits=c(0, 100))
```

:::
:::


