# . 3 Problem Set {#sec-P3_probSet .unnumbered}

::: callout-note
Please render this file to .html and submit via Brightspace.
:::

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(tidyverse)
library(readxl)
```

## Normal distribution

### Scenario 1: Using `*norm()`

For a particular population of brown seaweed, the winter carbon:nitrogen ratio is distributed normally among individuals as $y \sim Norm(\mu=27.6, \sigma=2.5)$. 

What proportion of the population would you expect to have a C:N ratio greater than 30?

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Proportion > 30"

mu <- 27.6
sigma <- 2.5

1 - pnorm(30, mu, sigma)
```
:::

What is the probability of sampling an individual with a ratio between 23 and 24?

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Pr(23 < y < 24)"

pnorm(24, mu, sigma) - pnorm(23, mu, sigma)
```
:::

What ratio would delineate the lowest 30% of the population?

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Lowest 30%"

qnorm(0.3, mu, sigma)
```
:::

Generate a random sample of 12 individuals from this population. Calculate the sample mean and sample standard deviation. 

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Collect a random sample"

obs <- rnorm(12, mu, sigma)
```

```{r}
#| code-fold: true
#| code-summary: "Sample mean and sd"
#| results: "hold"

# Mean:
mean(obs)

# sd:
sd(obs)
```

:::

What is the difference between your sample mean and the true population mean called? Why does it occur and what are two aspects of your sampling design that affect it?

> 

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Definition"
#| eval: false 

# This is sampling error: the difference between the true population parameter
# and the statistic calculated from a sample. It is inevitable that a statistic
# from a sample (e.g., a sample mean) will be different from the parameter for
# the entire population (e.g., the population mean) because a sample is only
# a subset of the whole population.
```

```{r}
#| code-fold: true
#| code-summary: "Sampling impact 1"
#| eval: false 

# Sampling error is strongly affected by:
# 1) Sample size: A larger sample will, on average, have smaller sampling error
#     since differences among individuals 'average out'
```

```{r}
#| code-fold: true
#| code-summary: "Sampling impact 2"
#| eval: false 

# Sampling error is strongly affected by:
# 2) Randomization: A non-random sample introduces bias, making some individuals
#     more likely to be included. This bias will tend to make sampling error 
#     biased as well.
```

```{r}
#| code-fold: true
#| code-summary: "Sampling impact 3"
#| eval: false 

# Sampling error is strongly affected by:
# 3) Population variability: A more variable population (larger sigma) means the
#     observations in a sample are more variable, and so sampling error will
#     tend to be larger. Standardizing conditions as much as possible can reduce
#     this source of noise.
```
:::


Repeat your random sampling 1000 times, calculate a sample mean for each of the 1000 samples, and make a histogram of them. Calculate the mean of these sample means. 

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Hint"
#| eval: false

# y_bar <- rep(0, 1000)
# for(i in 1:1000) {
#   y_bar[i] <- ...
# }
```

:::


::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Repeated sampling"

y_bar <- rep(0, 1000)
for(i in 1:1000) {
  y_bar[i] <- mean(rnorm(12, mu, sigma))
}
```

```{r}
#| code-fold: true
#| code-summary: "Histogram"
hist(y_bar, breaks=15); abline(v=mu, col="blue")
```

```{r}
#| code-fold: true
#| code-summary: "Mean of sample means"
#| results: "hold"
signif(mean(y_bar), 3)
```

How does this mean (and the histogram) compare to the population mean? 

>


```{r}
#| code-fold: true
#| code-summary: "Comparison sample means vs. mu"
#| eval: false

# The mean of sample means is very close to mu, and the histogram of sample means
# is centred on mu. 
#
# Note also that the distribution is approximately normal, illustrating the 
# Central Limit Theorem (since the population is ~Norm).
# 
# This relies on a *random* sample. If the sampling is biased (non-random), this
# pattern can not be relied upon!
```

:::




### Scenario 2: Transformations

The `swiss` dataset in R gives standardised fertility and socioeconomic indicators for each of the 47 French-speaking provinces of Switzerland in ca. 1888:

```{r}
str(swiss)
```

The values are given as percentages ranging from 0-100. Make a histogram for the `Education` column, setting the x-axis limits to the theoretical range of the variable.

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Histogram"

hist(swiss$Education, 
     xlim=c(0, 100), 
     main="Education beyond primary school",
     xlab="Percent of draftees, 1888")
```

:::

What transformation would be most appropriate for these kinds of data? 

>

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

# Logistic: Data are percentages (easily converted to proportions) and tend to be smaller values near the lower boundary (0%) with a long tail of larger values (right skew).

# A log transformation may also be suitable.
```

:::

Apply this transformation and make a histogram of the transformed values.

```{r}

```

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Transformation and histogram"

hist(boot::logit(swiss$Education/100), 
     main="Education beyond primary school",
     xlab="logit(Proportion of draftees), 1888")
```

:::

What do you notice about the shape of the original and transformed distributions?

>

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

# The transformed distribution looks much more like a normal distribution. 
```

:::


## Standard error of the mean

Calculations with populations vs. samples.



