# Problem Set 3 {#sec-P3_probSet .unnumbered}

::: callout-note

<a href="downloads/probSets_P3_student.qmd" download="probSets_P3_student.qmd">
  <i class="bi bi-download"></i>  Download file
</a> 

Please render this file to .html and submit via Brightspace.
:::

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(tidyverse); theme_set(theme_classic())
library(readxl)
set.seed(2026)
```

## Normal distribution

### Scenario 1: Using `*norm()`

For a particular population of brown seaweed, the winter carbon:nitrogen ratio is distributed approximately normally among individuals as $y \sim Norm(\mu=27.6, \sigma=2.5)$. 

What proportion of the population would you expect to have a C:N ratio greater than 30?

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Proportion > 30"

mu <- 27.6
sigma <- 2.5

1 - pnorm(30, mu, sigma)
```
:::

:::

What is the probability of sampling an individual with a ratio between 23 and 24?

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Pr(23 < y < 24)"

pnorm(24, mu, sigma) - pnorm(23, mu, sigma)
```
:::

:::

What ratio would delineate the lowest 30% of the population?

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Lowest 30%"

qnorm(0.3, mu, sigma)
```
:::

:::

Generate a random sample of 12 individuals from this population. Calculate the sample mean and sample standard deviation. 

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Collect a random sample"

obs <- rnorm(12, mu, sigma)
```

```{r}
#| code-fold: true
#| code-summary: "Sample mean and sd"
#| results: "hold"

# Mean:
mean(obs)

# sd:
sd(obs)
```

:::

:::

What is the difference between your sample mean and the true population mean called? Why does it occur and what are two aspects of your sampling design that affect it?

> 

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Definition"
#| eval: false 

# This is sampling error: the difference between the true population parameter
# and the statistic calculated from a sample. It is inevitable that a statistic
# from a sample (e.g., a sample mean) will be different from the parameter for
# the entire population (e.g., the population mean) because a sample is only
# a subset of the whole population.
```

```{r}
#| code-fold: true
#| code-summary: "Sampling impact 1"
#| eval: false 

# Sampling error is strongly affected by:
# 1) Sample size: A larger sample will, on average, have smaller sampling error
#     since differences among individuals 'average out'
```

```{r}
#| code-fold: true
#| code-summary: "Sampling impact 2"
#| eval: false 

# Sampling error is strongly affected by:
# 2) Randomization: A non-random sample introduces bias, making some individuals
#     more likely to be included. This bias will tend to make sampling error 
#     biased as well.
```

```{r}
#| code-fold: true
#| code-summary: "Sampling impact 3"
#| eval: false 

# Sampling error is strongly affected by:
# 3) Population variability: A more variable population (larger sigma) means the
#     observations in a sample are more variable, and so sampling error will
#     tend to be larger. Standardizing conditions as much as possible can reduce
#     this source of noise.
```
:::

:::


Repeat your random sampling 1000 times, calculate a sample mean for each of the 1000 samples, and make a histogram of them. Calculate the mean of these sample means. 

```{r}

```

::: callout-tip
```{r}
#| code-fold: true
#| code-summary: "Hint"
#| eval: false

# y_bar <- rep(0, 1000)
# for(i in 1:1000) {
#   y_bar[i] <- ...
# }
```

:::


::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Repeated sampling"

y_bar <- rep(0, 1000)
for(i in 1:1000) {
  y_bar[i] <- mean(rnorm(12, mu, sigma))
}
```

```{r}
#| code-fold: true
#| code-summary: "Histogram"
hist(y_bar, breaks=15); abline(v=mu, col="blue")
```

```{r}
#| code-fold: true
#| code-summary: "Mean of sample means"
#| results: "hold"
signif(mean(y_bar), 3)
```

:::

:::

How does this mean (and the histogram) compare to the population mean? 

>

::: {.content-hidden when-profile="man"}

::: callout-important

```{r}
#| code-fold: true
#| code-summary: "Comparison sample means vs. mu"
#| eval: false

# The mean of sample means is very close to mu, and the histogram of sample means
# is centred on mu. 
#
# Note also that the distribution is approximately normal, illustrating the 
# Central Limit Theorem (since the population is ~Norm).
# 
# This relies on a *random* sample. If the sampling is biased (non-random), this
# pattern can not be relied upon!
```

:::

:::




### Scenario 2: Transformations

The `swiss` dataset in R gives standardised fertility and socioeconomic indicators for each of the 47 French-speaking provinces of Switzerland in ca. 1888:

```{r}
str(swiss)
```

The values are given as percentages ranging from 0-100. Make a histogram for the `Education` column, setting the x-axis limits to the theoretical range of the variable.

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Histogram"

hist(swiss$Education, 
     xlim=c(0, 100), 
     main="Education beyond primary school",
     xlab="Percent of draftees, 1888")
```

:::

:::

What transformation would be most appropriate for these kinds of data? 

>

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

# Logistic: Data are percentages (convert to proportions by dividing by 100) and tend to be smaller values near the lower boundary (0%) with a long tail of larger values (right skew).

# A log transformation may also be suitable.
```

:::

:::

Apply this transformation and make a histogram of the transformed values.

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Transformation and histogram"

hist(boot::logit(swiss$Education/100), 
     main="Education beyond primary school",
     xlab="logit(Proportion of draftees), 1888")
```

:::

:::

What do you notice about the shape of the original and transformed distributions?

>

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

# The transformed distribution looks much more like a normal distribution. 
```

:::

:::

## Standard error of the mean

For the seaweed population in Scenario 1, we simulated a sample of 12 individuals from the population. In my particular sample, I found a sample mean of `{r} round(mean(obs), 3)`. Recall that the true population was distributed as $y \sim Norm(\mu=27.6, \sigma=2.5)$. 

Using the true population standard deviation $\sigma=2.5$, calculate the standard error of the mean. 

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"

sigma <- 2.5
n <- 12

sigma / sqrt(n)
```

:::

:::

In your own words, describe what the standard error of the mean is. 

>

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution (my words)"
#| eval: false

# The standard error of the mean is the standard deviation of the distribution of possible sample means from a given population. It describes the breadth of the distribution of sample means we would expect if we were to collect many, many different samples, each with size n, and calculate a mean from each sample.
```

:::

:::

The distribution of sample means can be simulated in R. Write code to 1) simulate 100 different samples from the seaweed population with $n=12$, and 2) calculate a sample mean from each of these 100 samples.

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"

n_samples <- 100
n <- 12
sample_means <- rep(0, n_samples)
for(i in 1:n_samples) {
  sample_means[i] <- rnorm(n, mu, sigma) |> mean()
}
```

:::

:::


Make a plot to show the distribution of these sample means, and indicate the true population mean $\mu$.

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"

hist(sample_means, breaks=10)
abline(v=mu, col="red", lwd=2)
```

:::

:::

Calculate the standard deviation of the 100 sample means. 

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"

sd(sample_means)
```

:::

:::

How does it compare to the standard error of the mean you calculated above? Try increasing the number of samples to 3,000 and recalculate.

>

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

# The standard deviation of the sample means should be close. Increasing the number of samples should bring this value even closer to the standard error of the mean calculated analytically. Note that this is because we are averaging over sampling error! 
```

:::

:::


Lastly, repeat the process, this time with $n=120$ for each sample. Calculate the standard error of the mean, simulate 100 sample means, and compare the standard deviation of these with the calculated standard error.

```{r}

```

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"

n <- 120

sigma / sqrt(n)
n_samples <- 100
sample_means <- rep(0, n_samples)
for(i in 1:n_samples) {
  sample_means[i] <- rnorm(n, mu, sigma) |> mean()
}
hist(sample_means, breaks=10)
abline(v=mu, col="red", lwd=2)
sd(sample_means)
```

:::

:::

Why does the standard error decrease? What does this tell us about our level of confidence in the sample statistics we calculate ($\bar{y}$, $s$) and our inferences about the population parameters we are interested in ($\mu$, $\sigma$)?

>

::: {.content-hidden when-profile="man"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Solution"
#| eval: false

# Sample means will vary less (among samples) as the size of each sample increases. The variability among individuals averages out, and our samples become more representative of the population. According to the equation, the standard error decreases with the square root of the sample size. 

# So, with a larger sample size, we can be more confident that our sample mean is near to the population mean. 

# Note that this assumes proper sampling methodology! (See Experimental Design lectures)
```

:::

:::
