# Problem Set 1 {#sec-P1_probSet .unnumbered}

::: callout-note
Please render this file to .html and submit via Brightspace.
:::

```{r}
#| message: false
#| warning: false
#| code-fold: false

library(tidyverse)
library(readxl)
set.seed(1)
```

## R Basics

### Data types

Create a vector called `site_num` with the values 1 through 10.

```{r}

```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Three methods which create identical vectors"

site_num <- 1:10
site_num <- seq(1, 10, by=1)
site_num <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
```
:::

:::

## Data munging

The *SmoltWeights* sheet from H2SMS_practicalData.xlsx contains the weight in grams (column `weight_g`) from a sample of 34 wild salmon smolts (column `fish_id`) during their out-migration from a river to the ocean. Load the dataset and perform basic summaries (range, mean, median, number of missing values for each column). Identify (using code) which fish are missing values for `weight_g`.

```{r}

```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Load and check data"

smolt_df <- read_xlsx("data/H2SMS_practicalData.xlsx", sheet="SmoltWeights")
str(smolt_df)
summary(smolt_df)
```

There are many ways to identify the row with missing data.

```{r}
#| code-fold: true
#| code-summary: "Method 1: Identify the row number"

which(is.na(smolt_df$weight_g))
```

```{r}
#| code-fold: true
#| code-summary: "Method 2: Extract the row from the data.frame"

smolt_df[is.na(smolt_df$weight_g),] # gives the row from the data.frame
```

```{r}
#| code-fold: true
#| code-summary: "Method 3: Filter the data.frame"

smolt_df |>
  filter(is.na(weight_g))
```
:::

:::

Create a new column called `weight_lb` that gives the weights in pounds, where 1 kg = 2.205 lb.

```{r}

```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Calculate new column"

smolt_df$weight_lb <- smolt_df$weight_g / 1000 * 05
```
:::

:::

What proportion of fish weigh less than 0.2 lb? How many fish is this?

```{r}

```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Proportion"

mean(smolt_df$weight_lb < 0.2, na.rm=T)
```

```{r}
#| code-fold: true
#| code-summary: "Number"

sum(smolt_df$weight_lb < 0.2, na.rm=T)
```
:::

:::

Create a new data.frame called `large_df` that contains only the fish as large or larger than the median weight *and* excludes rows with missing weights.

```{r}

```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| echo: false
smolt_df |>
  drop_na(weight_g) |>
  filter(weight_g >= median(weight_g))
```

```{r}
#| code-fold: true
#| code-summary: "Using base R functions"

smolt_median <- median(smolt_df$weight_g, na.rm=T)
large_df <- smolt_df[smolt_df$weight_g >= smolt_median & !is.na(smolt_df$weight_g), ]
```

```{r}
#| code-fold: true
#| code-summary: "Using tidyverse functions"

large_df <- smolt_df |>
  drop_na(weight_g) |>
  filter(weight_g >= median(weight_g))
```
:::

:::

Create a new data.frame called `extreme_df` that contains only fish more than 1.5 standard deviations greater than the mean *and* excludes rows with missing weights.

```{r}

```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| echo: false

smolt_df |>
  drop_na(weight_g) |>
  filter(weight_g > mean(weight_g) + 1.5*sd(weight_g) |
           weight_g < mean(weight_g) - 1.5*sd(weight_g))
```

```{r}
#| code-fold: true
#| code-summary: "Using base R functions"

smolt_mean <- mean(smolt_df$weight_g, na.rm=T)
smolt_sd <- sd(smolt_df$weight_g, na.rm=T)
extreme_df <- smolt_df[!is.na(smolt_df$weight_g) &
                         (smolt_df$weight_g > smolt_mean + 1.5*smolt_sd |
                            smolt_df$weight_g < smolt_mean - 1.5*smolt_sd), ]
```

```{r}
#| code-fold: true
#| code-summary: "Using tidyverse functions"

extreme_df <- smolt_df |>
  drop_na(weight_g) |>
  filter(weight_g > mean(weight_g) + 1.5*sd(weight_g) |
           weight_g < mean(weight_g) - 1.5*sd(weight_g))
```
:::

:::

## Plotting

### Histograms

Create a histogram of the smolt weights, labelling the axes and omitting a plot title. Add a figure legend using `#| fig-cap: "..."` in the quarto document code block. Add `#| label: fig-smoltHist`, and then in *text* below, reference the figure by typing `@fig-smoltHist`.

::: callout-tip
When rendered, `@fig-smoltHist` in the text is automatically replaced by "Figure 1", with figure numbers updated automatically. Adding a legend with `#| fig-cap: "..."` in your code block will automatically give that figure a number, making your life easier.

The label must start with `fig-` for this to work.

Note that you *cannot* re-use the same `label`.
:::

```{r}
#| fig-cap: "..."


```

::: {.content-visible when-profile="key"}

::: callout-important
Here is some text referencing @fig-smoltHist by using the label I've used in the code chunk (`@fig-smoltHist`).

```{r}
#| code-fold: true
#| code-summary: "Code chunk"
#| echo: fenced
#| label: fig-smoltHist
#| fig-cap: "Histogram of smolt weights."
#| fig-width: 4
#| fig-height: 4

hist(smolt_df$weight_g, 
     main=NULL, 
     xlab="Smolt weight (g)",
     ylab="Number of fish")
```
:::

:::

### Line plots

The dataset `beaver1` is available in R, giving the body temperature through time of a female beaver (*Castor canadensis*), with observations every 10 minutes. 

```{r}
str(beaver1)
```

Debug the following code so that it produces a line plot of body temperature by observation id. Note that `time` is given in the form `0330` for 3:30am (`?beaver1`). The `obs_id` column thus gives the ordered observations from first to last. 

As an added challenge, add vertical dashed lines indicating when the beaver was active.

::: callout-tip
Set `#| eval: true` to run the code when you click Render. If you've successfully debugged, you should then see your plot.
:::

```{r}
#| eval: false

beaver_df <- beaver1
beaver_df$obs_id <- 1:nrow(beaver_df)
plot(obs_id ~ Temp, 
     data=beaver_df, 
     type="p", 
     xLab="Observation (time)",
     ylab="Body temperature (C)")
```

::: {.content-visible when-profile="key"}

::: callout-important

Original:
```{r}
#| eval: false

beaver_df <- beaver1
beaver_df$obs_id <- 1:nrow(beaver_df)
plot(obs_id ~ Temp, # Order is flipped (y ~ x) and `temp` should be lower case
     data=beaver_df, 
     type="p", # Should be "l" (line) or "b" (line with points) 
     xLab="Observation (time)", # typo: lower case l
     ylab="Body temperature (C)")
```

Corrected:
```{r}
#| eval: true

beaver_df <- beaver1
beaver_df$obs_id <- 1:nrow(beaver_df)
plot(temp ~ obs_id, data=beaver_df, type="l", 
     xlab="Observation (time)",
     ylab="Body temperature (C)")
# ?abline
abline(v=beaver_df$obs_id[beaver_df$activ==1], lty=2)
```
:::

:::


### Boxplots

The dataset `ToothGrowth` is available in R, giving the length (`len`) of tooth-growth cells (odontoblasts) in 60 guinea pigs under different vitamin C regimes (`supp`: OJ=orange juice, VC=ascorbic acid; `dose`: 0.5, 1, 2 mg/day). Create a boxplot of the length by dose for only the guinea pigs receiving orange juice. Write a figure caption and add a label as above.

```{r}
#| fig-cap: "..."


```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Using base R"
#| echo: fenced
#| label: fig-OJ_lenByDose
#| fig-cap: "Odontoblast length by dosage in guinea pigs receiving orange juice."
#| fig-width: 5
#| fig-height: 3.5

boxplot(len ~ dose, data=ToothGrowth[ToothGrowth$supp=="OJ",],
        xlab="Dose (mg/day)",
        ylab="Odontoblast length (yards?)")
```
:::

:::

Use `ggplot()` to make the same boxplot, but with `supp` as the fill colour.

::: callout-tip
Remember to check the column types first!
:::

```{r}
#| fig-cap: "..."


```

::: {.content-visible when-profile="key"}

::: callout-important
```{r}
#| code-fold: true
#| code-summary: "Check data.frame structure"

str(ToothGrowth)
```

Since `dose` is numeric, there are two options to produce different boxes for each combination of `dose` and `supp`.

```{r}
#| code-fold: true
#| code-summary: "Method 1: Coerce `dose` to a factor"
#| echo: fenced
#| label: fig-lenByDose_gg1
#| fig-cap: "Odontoblast length by dosage and supplement type in guinea pigs."
#| fig-width: 5
#| fig-height: 3.5

ggplot(ToothGrowth, aes(factor(dose), len, fill=supp)) + 
  geom_boxplot() +
  scale_fill_manual("Supplement", values=c("orange", "steelblue")) +
  labs(x="Dose (mg/day)", 
       y="Odontoblast length (nm)")
```

```{r}
#| code-fold: true
#| code-summary: "Method 2: Specify `group=`"
#| echo: fenced
#| label: fig-lenByDose_gg2
#| fig-cap: "Odontoblast length by dosage and supplement type in guinea pigs."
#| fig-width: 5
#| fig-height: 3.5

ggplot(ToothGrowth, aes(dose, len, fill=supp, group=paste(dose, supp))) + 
  geom_boxplot() +
  scale_fill_manual("Supplement", values=c("orange", "steelblue")) +
  labs(x="Dose (mg/day)", 
       y="Odontoblast length (nm)")
```

::: callout-note
What are the benefits and downsides of each method? When might you choose one or the other?
:::

:::

:::

### Scatter plots

Using the same `ToothGrowth` dataset, produce a scatter plot of odontoblast length as a function of dose, where `supp` is the point colour.

```{r}
#| fig-cap: "..."


```

::: {.content-visible when-profile="key"}

::: callout-important
Since `supp` is a factor, the *levels* can be coded into colours automatically in `plot()`.

```{r}
#| code-fold: true
#| code-summary: "Using base R"
#| echo: fenced
#| label: fig-lenByDose_baseRScatter
#| fig-cap: "Odontoblast length by dosage and supplement type in guinea pigs."
#| fig-width: 4
#| fig-height: 4

plot(len ~ dose, data=ToothGrowth, col=supp,
     xlab="Dose (mg/day)",
     ylab="Odontoblast growth (AU)")
legend("topleft", 
       legend=levels(ToothGrowth$supp), 
       col=1:2, 
       pch=1, 
       bty="n")
```

In my opinion, it is easier to use ggplot as your figures get more complex.

```{r}
#| code-fold: true
#| code-summary: "Using ggplot"
#| echo: fenced
#| label: fig-lenByDose_ggScatter
#| fig-cap: "Odontoblast length by dosage and supplement type in guinea pigs."
#| fig-width: 4.5
#| fig-height: 3

ggplot(ToothGrowth, aes(dose, len, colour=supp)) + 
  geom_point(shape=1, size=2, stroke=0.8) +
  scale_colour_manual("Supplement", values=c("orange", "steelblue")) +
  labs(x="Dose (mg/day)", 
       y="Odontoblast length (nm)") +
  theme_classic()
```
:::

:::
