---
title: "Extra R: CA1"
subtitle: "H2 Statistics for Marine Science"
author: "Name"
date: today
format:
  html:
    theme: lux
    code-block-bg: true
    code-link: true
    toc: true
    number-sections: true
    callout-appearance: simple
    embed-resources: true
editor: visual
---

# Extra: CA1

## Overview 

Coverage: L01-L10, P1-P3

This is a document to give you additional practice working in R. It will include working with data (manipulating, summarising, plotting) and analyzing the distributions we've covered.

Please do not treat this as a comprehensive study guide for CA1!

Note that there are often many possible paths to the correct solution.

## Iris

The `iris` dataset is available in R upon start-up.

What are the dimensions of the object (rows, columns) and what data type is it?

```{r}

```


Create a scatter plot of sepal area on the x-axis and petal area on the y-axis.

```{r}

```


For each species, what is the mean and standard deviation of sepal area?

```{r}

```



Create a plot with two panels, each with a boxplot with species on the x-axis. In the first panel, plot sepal area on the y-axis. In the second panel, plot petal area on the y-axis.

```{r}

```



Extract the even numbered rows from `iris`, and the columns `Species`, `Sepal.Area`, and `Petal.Area` in that order.

```{r}

```



## Air quality

The `airquality` dataset is available in R upon start-up, and gives daily values of mean ozone, solar radiation, mean wind speed, and max temperature at LaGuardia Airport in New York City (`?airquality`).

Make a histogram of wind speed.

```{r}

```


Assume wind speed is normally distributed in this dataset with $\mu=9.99 mph$ and $\sigma = 3.51 mph$.

What proportion of days would you expect to have a wind speed less than 5 mph?

```{r}

```


You have a kite that can only be flown safely in winds between 14 mph and 20 mph. What proportion of days would be suitable?

```{r}

```


How many days would be suitable in a 31-day month?

```{r}

```


What proportion of days in the dataset were actually suitable?

```{r}

```


What are the z-scores for 14 and 20 mph?

```{r}

```


What are the quantiles for the middle 95%?

```{r}

```


What is the quantile for the largest 5%?

```{r}

```


Given the population parameters above, what is the z score of the sample mean in the dataset?

```{r}

```


Given the population parameters above, what is the probability of observing a sample mean smaller than the one in this dataset?

```{r}

```


Create a scatter plot with temperature on the x-axis and log-transformed ozone on the y-axis.

```{r}

```


Create a boxplot with the solar radiation on the y-axis and the month on the x-axis.

```{r}

```


## Coral predation

You are studying predation of the Flamingo Tongue Snail on soft corals on South Water Caye in Belize. In a distinct patch of reef, you randomly select a sample of 14 corals and identify that 9 of these have feeding scars.

What distribution would you expect this variable (i.e., the number of corals with feeding scars of those sampled) to follow?

>

What is your sample proportion and its standard error?

```{r}

```


Create a bar graph of the binomial distribution corresponding to these parameters, with the x-axis spanning the full range of possible outcomes.

```{r}

```


Based on this distribution, if you were to collect another random sample of the same size, what is the probability that all 14 would have feeding scars?

```{r}

```


What is the probability you would find *fewer* with feeding scars than in the first sample?

```{r}

```


What is the probability of finding *more* with feeding scars than in the first sample?

```{r}

```

What is the probability of a sample with 4, 5, or 6 corals with feeding scars?

```{r}

```


## Earthquakes

The dataset `quakes` is available in R upon start-up, and gives magnitudes and locations for 1000 seismic events near Fiji since 1964.

Plot the locations of the earthquakes (x-axis: longitude, y-axis: latitude).

```{r}

```


Create a column called `mag_cat` with the magnitude truncated to integer (i.e., 4.0-4.99 become 4). Hint: `?trunc`

```{r}

```


For each magnitude category (`mag_cat`), count the number of occurrences in the dataset.

```{r}

```


Let's assume the dataset spans 60 years (1964-2023). What is the mean number of magnitude category 5 earthquakes per year?

```{r}

```


What distribution is most suitable for the number of category 5 earthquakes per year?

>


Using the above value as $\lambda$, create a bar plot of the probability distribution of the number of category 5 earthquakes per year, ranging from 0 to 10 earthquakes.

```{r}

```


What is the probability that next year has no category 5 earthquakes?

```{r}

```


Of the next 10 years, how many would you expect to have:

No category 5 earthquakes?

```{r}

```


Fewer than 2 earthquakes?

```{r}

```


At least 5?

```{r}

```


